Some examples in extended logic counting classes

### 2016-12-28 LICS/appendix-notation.tex

%!TEX root = main.tex

\subsection{Notation for the appendix}

@@

If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$. We also use $\length(\bar{x})$ ($\length(\bar{X})$) to denote the number or variables in the tuple.
%\subsection{Proof of Theorem \ref{no-mult}}
%
%Let $\{+, \cdot, \sa{}, \pa{}\}$ be the set of {\it algebraic operators} in $\qso$. We will prove by induction over the number of algebraic operators in a $\qso$-formula $\alpha$ that there exists some $\qso$-formula $\alpha'$ that respects the stated condition. Let $\text{op}(\alpha)$ be the number of algebraic operators in $\alpha$. The case $\text{op}(\alpha) = 0$ is trivial. Let $n\in\nat$ be such that for each $\qso$-formula $\beta$ such that $\text{op}(\beta) < n$ there is some $\beta'$ that respects the condition. Let $\alpha$ be a $\qso$-formula such that $\text{op}(\alpha) = n$. Note that if $\alpha$ is of the form $\alpha = (\beta_1 \add \beta_2)$, $\alpha = \sa{x} \beta$, $\alpha = \pa{x} \beta$, $\alpha = \sa{X} \beta$ or $\alpha = \pa{X} \beta$ the proof follows directly. Therefore we assume that $\alpha = (\beta_1 \mult \beta_2)$. We separate the proof in two main cases:
%\begin{enumerate}
%	\item At least one of $\beta_1, \beta_2$ is not of the form $\pa{x} \gamma$, $\pa{X} \gamma$, or $\varphi$, where $\varphi$ is an $\so$-formula. Without loss of generality, let $\beta_1$ be such one.
%	\begin{enumerate}
%		\item $\beta_1 = s \in \nat$. Then we use $\alpha' = \beta_2 \add \cdots \add \beta_2$, where $\beta_2$ is repeated $s$ times, and if $s = 0$, take $\alpha' = \perp$. Since each of the formulas paired by $\add$ has less than $n$ algebraic operators, the statement holds.
%		\item $\beta_1 = (\gamma_1 \add \gamma_2)$. Then, we use $\alpha' = (\gamma_1\mult\beta_2 + \gamma_2\mult\beta_2)$ and each of the formulas paired by $\add$ have less than $n$ algebraic operators so the statement holds.
%		\item $\beta_1 = \sa{x} \gamma$. Then, we use $\alpha' = \sa{x} (\gamma\mult\beta_2)$ and as in the previous case the statement holds.
%		\item $\beta_1 = \sa{X} \gamma$, which we treat exactly as the previous one.
%	\end{enumerate}
%	\item Both of $\beta_1$ and $\beta_2$ are of the form $\pa{x} \gamma$, $\pa{X} \gamma$, or $\varphi$, where $\varphi$ is an $\so$-formula. This case also separates in further cases.
%	\begin{enumerate}
%		\item $\beta_1 = \pa{x} \gamma_1(x)$ and $\beta_2 = \pa{y} \gamma_2(y)$. Then we use $\alpha' = \pa{x}(\gamma_1(x)\mult\gamma_2(x))$, and the statement holds.
%		\item $\beta_1 = \pa{X} \gamma_1(X)$ and $\beta_2 = \pa{Y} \gamma_2(Y)$. [something about arities].
%		\item $\beta_1 = \pa{X} \gamma_1(X)$ and $\beta_2 = \pa{x} \gamma_2(x)$ (w.l.o.g.). [use singleton predicates].
%		\item $\beta_1 = \pa{x} \gamma$ and $\beta_2 = \varphi$, which is an $\so$-formula (w.l.o.g.). Then, we use $\alpha' = \pa{x} (\gamma \cdot \varphi)$, and the statement holds.
%		\item $\beta_1 = \pa{X} \gamma$ and $\beta_2 = \varphi$, which is an $\so$-formula (w.l.o.g.). We treat this case exactly as the previous one.
%		\item $\beta_1 = \varphi_1$ and $\beta_2 = \varphi_2$ both $\so$-formulas. Then, we use $\alpha' = (\varphi_1 \wedge \varphi_2)$ and the statement holds.
%	\end{enumerate}
%\end{enumerate}

\subsection{Proof of Theorem \ref{captfp}}

For the first condition, let $\alpha\in\qfo(\lfp)$ over some signature $\R$. Let $f$ be a function over $\R$ defined by the following procedure. Let $\enc(\A)$ be an input, where $\A$ is an ordered structure over $\R$ with domain $A = \{a_1,\ldots,a_n\}$. Replace each first order sum and first order product in $\alpha$ by an expansion using the elements in $A$. This is, $\Sigma x \beta(x)$ is replaced by $(\beta(a_1)+\cdots+\beta(a_n))$ and $\Pi x \beta(x)$ is replaced by $(\beta(a_1)\cdot\,\cdots\,\cdot\beta(a_n))$. Then each sub-formula $\varphi\in\lfp$ in $\alpha$ is replaced by 1 if $\A\models\varphi$ and by 0 otherwise. The resulting formula is an arithmetic expression of polynomial size which is evaluated and lastly given as output. Note that $f\in\fp$ and $f(\enc(\A)) = \alpha(\enc(\A))$.

For the second condition, let $f\in \fp$ defined over some signature $\R$. Let $\ell\in\nat$ be such that for each $\A\in\ostr[\R]$, $\lceil\log_2 f(\enc(\A)) \rceil \leq n^\ell$, where $\A$ has a domain of size $n$. Let $\Phi(x_0,\ldots,x_{\ell-1})$ be a formula that represents the following. Let $\A$ be an ordered structure over $\R$ with domain $A$. Consider a linear order over $A^{\ell}$ given by $\bar{a} < \bar{b}$, for any $\bar{a},\bar{b}\in A^{\ell}$. Let $w$ be a string of size $n^{\ell}$ the represents $f(\enc(\A))$ in binary. Let $(a_1,\ldots,a_{\ell})\in A^{\ell}$ be the $m$-th tuple in the given order (starting from 0). $\A\models\Phi(a_1,\ldots,a_{\ell})$ if an only if the $m$-th bit in $w$ is 1, from least to most significant. Since this condition can be computed in polynomial time, and $\lfp$ captures $\ptime$, this formula can be expressed in $\lfp$. We use
$$
\alpha = \sa{x_1} \cdots \sa{x_{\ell}} \Phi(x_1,\ldots,x_{\ell})\cdot\varphi(x_1,\ldots,x_{\ell}),
$$
where $\varphi(x_1,\ldots,x_{\ell}) := \pa{y_1}\cdots\pa{y_{\ell}}((y_1,\ldots,y_{\ell}) < (x_1,\ldots,x_{\ell}) \mapsto 2).$ Note that if $\bar{a} \in A^{\ell}$ is the $m$-th tuple in the given order (starting from 0), then $\sem{\varphi(\bar{a})}(\A) = 2^{m}$. Adding this values for each $\bar{a}\in A^{\ell}$ gives exactly $f(\enc(\A))$. Then, $\alpha$ is in $\qfo(\lfp)$ over $\R$ and $\alpha(\enc(\A)) = f(\enc(\A))$. \qed

\subsection{Proof of Theorem \ref{qfo-pfp-cap}}

\martin{no revisar, es un copy-paste de arriba}

For condition (2) in \ref{cap-def}, let $\alpha\in\qfo(\pfp)$ over some $\R$. Let $f$ be a function over $\R$ defined by the following procedure. Let $\enc(\A)$ be an input, where $\A$ is an ordered structure over $\R$ with domain $A = \{a_1,\ldots,a_n\}$. Replace each first order sum and first order product in $\alpha$ by an expansion using the elements in $A$. This is, $\Sigma x \beta(x)$ is replaced by $(\beta(a_1)+\cdots+\beta(a_n))$ and $\Pi x \beta(x)$ is replaced by $(\beta(a_1)\cdot\,\cdots\,\cdot\beta(a_n))$. Then each sub-formula $\varphi\in\pfp$ in $\alpha$ is replaced by 1 if $\A\models\varphi$ and by 0 otherwise. The resulting formula is an arithmetic expression of polynomial size which is evaluated and lastly given as output. Note that $f\in\nfpspace$ and $f(\enc(\A)) = \alpha(\enc(\A))$.

For condition (2) in \ref{cap-def}, let $f\in \nfpspace$ defined over some $\R$. Let $\ell\in\nat$ be such that for each $\A\in\ostr[\R]$, $\lceil\log_2 f(\enc(\A)) \rceil \leq n^\ell$, where $\A$ has a domain of size $n$. Let $\Phi(x_0,\ldots,x_{\ell-1})$ be a $\pfp$ formula that represents the following. For each $\A \in \ostr[\R]$, for convenience, suppose its domain is $A = \{0,\ldots,n-1\}$ with the usual order. Let $y$ be a string of size $n^{\ell}$ that represents the value of $f(\enc(\A))$ in binary. Then, for each $\ell$-tuple $(a_0,\ldots,a_{\ell-1}) \in A^{\ell}$, it holds that $\A \models \Phi(a_0,\ldots,a_{\ell-1})$ if and only if the $(n^{\ell-1}a_{\ell-1} + \cdots + n^2 a_2 + n a_1 + a_0)$-th bit of $y$, from least to most significant, is 1. We use
$$
\alpha = \Sigma x_0 \cdots \Sigma x_{\ell-1} \Phi(x_0,\ldots,x_{\ell-1})\cdot\varphi_{\ell-1}(x_{\ell-1})\cdot\,\cdots\,\cdot\varphi_0(x_0).
$$
where $\varphi_i(x) = \Pi y[(y < x)\mapsto\Pi z_1\cdots\Pi z_i\,2]$ for $i > 0$ and $\varphi_0(x) = \Pi y[(y < x)\mapsto 2]$. Let $\A$ the previously defined structure. Note that for each $a\in A$, then $\sem{\varphi_i(x)}(\A,a) = 2^{n^i a}$. Therefore, for each $(a_0,\ldots,a_{\ell-1})\in A^\ell$, we have that
$$
\sem{\Phi(x_0,\ldots,x_{\ell-1})\cdot\varphi_{\ell-1}(x_{\ell-1})\,\cdots\,\varphi_0(x_0)}(\A,a_0,\ldots,a_{\ell-1}) = 
\begin{cases}
2^{(n^{\ell-1}a_{\ell-1} + \cdots + n^2 a_2 + n a_1 + a_0)} &\text{if } (\A,a_0,\ldots,a_{\ell-1})\models\Phi \\
0 &\text{otherwise.}
\end{cases}
$$
and adding these values gives $f(\enc(\A))$. Then, $\alpha$ is in $\qfo(\pfp)$ over $\R$ and $\alpha(\enc(\A)) = f(\enc(\A))$. \qed

\subsection{Proof of Theorem \ref{eqso-fo-cap}}

For condition (1) in \ref{cap-def}, we will construct a recursive non-deterministic algorithm $M_{\alpha}$ for each $\eqso(\fo)$ formula $\alpha$ over a signature $\R$. This machine, on input $(\A,v,V)$ accepts in $\sem{\alpha}(\A,v,V)$ of its non-deterministic paths for each $(\A,v,V) \in \ostr[\R]^*$. Suppose $\A$ has domain $A$. If $\alpha$ is a $\fo$-formula $\varphi$, then the algorithm checks if $(\A,v,V)\models\varphi$ deterministically in polynomial time, and accepts if and only if it holds true. If $\alpha$ is a constant $s$, it produces $s$ branches and accepts in all of them. If $\alpha = (\beta \add \gamma)$, then it chooses between 0 or 1, if it is 0 (1), it simulates $M_{\beta}$ ($M_{\gamma}$) on input $(\A,v,V)$. If $\alpha = (\beta \mult \gamma)$, it simulates $M_{\beta}$ on input $(\A,v,V)$ and on each accepting path, it continues simulating $M_{\gamma}$ on input $(\A,v,V)$.%%If $\alpha = \sa{x}\beta$, it chooses $a\in A$ and simulates $M_{\beta}$ on input $(\A,v[a/x],V)$.
If $\alpha = \pa{x}\beta$, it simulates $M_{\beta}$ on input $(\A,v[a/x],V)$ consecutively for each $a\in A$. If $\alpha = \sa{X}\beta$, it chooses $B\in A^{arity(X)}$ and simulates $M_{\beta}$ on input $(\A,v,V[B/X])$. This covers all possible cases for $\alpha$. Let $\alpha$ be a formula in $\eqso(\fo)$ over a signature $\R$ and let $f$ be a function over $\R$ such that $f(\enc(\A))$ is equal to the accepting paths of $M_{\alpha}$ on input $(\A,v,V)$ for some $(\A,v,V) \in \ostr[\R]^*$. We have that $f$ is a $\shp$ function over $\R$ and that for every $\A\in\ostr[\R]$: $f(\enc(\A)) = \alpha(\enc(\A))$.

For condition (2) in \ref{cap-def}, note that Saluja et al. \cite{SalujaST95} proved that $\shp = \#\fo$. We also have that $\sqso(\fo)$ captures $\#\fo$ over ordered structures so for each $f\in \shp$ let $\alpha \in \sqso(\fo)$ be its corresponding formula. Since this formula is also in $\eqso(\fo)$, the condition holds. \qed

\subsection{Proof of Theorem \ref{eqso-eso-cap}}

For condition (1) in \ref{cap-def}, we will construct a recursive non-deterministic algorithm $M_{\alpha}$ for each $\eqso(\eso)$ formula $\alpha$ over a signature $\R$. This machine, on input $(\A,v,V)$, accepts in $\sem{\alpha}(\A,v,V)$ of its non-deterministic outputs for each $(\A,v,V) \in \ostr[\R]^*$. Suppose $\A$ has domain $A$. On each of the steps the algorithm adds the current configuration to the output tape. If $\alpha$ is a $\eso$-formula $\varphi$ it checks if $(\A,v,V)\models\varphi$ non-deterministically in polynomial time without modifying the output tape, and accepts if and only if the condition holds true. If $\alpha$ is a constant $s$, then the algorithm produces $s$ branches and accepts in all of them. If $\alpha = (\beta \add \gamma)$, then it chooses between 0 or 1, if it is 0 (1), it simulates $M_{\beta}$ ($M_{\gamma}$) on input $(\A,v,V)$. If $\alpha = (\beta \mult \gamma)$, it simulates $M_{\beta}$ on input $(\A,v,V)$ and on each accepting path, it continues simulating $M_{\gamma}$ on input $(\A,v,V)$. %% If $\alpha = \sa{x}\beta$, it chooses $a\in A$ and simulates $M_{\beta}$ on input $(\A,v[a/x],V)$.
If $\alpha = \pa{x}\beta$, it simulates $M_{\beta}$ on input $(\A,v[a/x],V)$ consecutively for each $a\in A$. If $\alpha = \sa{X}\beta$, it chooses $B\in A^{arity(X)}$ and simulates $M_{\beta}$ on input $(\A,v,V[B/X])$. This covers all possible cases for $\alpha$. Let $\alpha$ be a formula in $\eqso(\eso)$ over a signature $\R$ and let $f$ be a function over $\R$ such that $f(\enc(\A))$ is equal to the number of accepting outputs of $M_{\alpha}$ on input $(\A,v,V)$ for some $(\A,v,V) \in \ostr[\R]^*$. We have that $f$ is a $\spp$ function over $\R$ and that for every $\A\in\ostr[\R]$: $f(\enc(\A)) = \alpha(\enc(\A))$.

For condition (2) in \ref{cap-def}, note that Compton and Graedel \cite{ComptonG96} proved that $\spp = \#\eso$. We also have that $\sqso(\eso)$ captures $\#\eso$ over ordered structures so for each $f\in \spp$ let $\alpha \in \sqso(\eso)$ be its corresponding formula. Since this formula is also in $\eqso(\eso)$, the condition holds. \qed

\subsection{Proof of Theorem \ref{qso-pfp-cap}}

For condition (1) in \ref{cap-def}, we will construct a recursive non-deterministic algorithm $M_{\alpha}$ for each $\qso(\pfp)$ formula $\alpha$ over a signature $\R$. This machine, on input $(\A,v,V)$, accepts in $\sem{\alpha}(\A,v,V)$ of its non-deterministic paths for each $(\A,v,V) \in \ostr[\R]^*$. Suppose $\A$ has domain $A$. If $\alpha$ is a $\pfp$-formula $\varphi$, then the algorithm checks if $(\A,v,V)\models\varphi$ deterministically in polynomial space \cite{AbiteboulV89}\martin{no estoy seguro de esta cita}, and accepts if and only if it holds true. If $\alpha$ is a constant $s$, it produces $s$ branches and accepts in all of them. If $\alpha = (\beta \add \gamma)$, then it chooses between 0 or 1, if it is 0 (1), it simulates $M_{\beta}$ ($M_{\gamma}$) on input $(\A,v,V)$. If $\alpha = (\beta \mult \gamma)$, it simulates $M_{\beta}$ on input $(\A,v,V)$ and on each accepting path, it continues simulating $M_{\gamma}$ on input $(\A,v,V)$. If $\alpha = \sa{x}\beta$, it chooses $a\in A$ and simulates $M_{\beta}$ on input $(\A,v[a/x],V)$. If $\alpha = \pa{x}\beta$, it simulates $M_{\beta}$ on input $(\A,v[a/x],V)$ consecutively for each $a\in A$. If $\alpha = \sa{X}\beta$, it chooses $B\in A^{arity(X)}$ and simulates $M_{\beta}$ on input $(\A,v,V[B/X])$. If $\alpha = \pa{X}\beta$, it simulates $M_{\beta}$ on input $(\A,v,V[B/X])$ consecutively for each $B\in A^{arity(X)}$. This covers all possible cases for $\alpha$, and each of these steps can be computed in polynomial space. Let $\alpha$ be a formula in $\qso(\pfp)$ over a signature $\R$ and let $f$ be a function over $\R$ such that $f(\enc(\A))$ is equal to the accepting paths of $M_{\alpha}$ on input $(\A,v,V)$ for some $(\A,v,V) \in \ostr[\R]^*$. We have that $f$ is a $\shpspace$ function over $\R$, which implies that $f$ is also a $\fpspace$ function over $\R$ \cite{Ladner89}, and that for every $\A\in\ostr[\R]$: $f(\enc(\A)) = \alpha(\enc(\A))$.

For the second condition, let $f\in \fpspace$ defined over some $\R$. Let $\ell\in\nat$ be such that for each $\A\in\ostr[\R]$, $\lceil\log_2 f(\enc(\A)) \rceil \leq 2^{n^\ell}$, where $\A$ has a domain of size $n$. Let $\Phi(X)$ be a $\pfp$ formula that represents the following. Consider an order over the subsets of $A^{\ell}$ given by the formula $$\varphi_{<}(X,Y) = \exists\bar{u}\big[\neg X(\bar{u})\wedge Y(\bar{u})\wedge \forall\bar{v}\big(
\bar{u}<\bar{v}\to(X(\bar{u})\iff Y(\bar{v}))\big)\big].$$

Let $B$ be the $p$-th subset with respect to this order and let $w$ be a string of size $2^{n^{\ell}}$ which represents the value of $f(\enc(\A))$ in binary. Then $\A\models\Phi(B)$ if and only if the $p$-th bit in $w$ is 1, from least to most significant. Since this condition can be verified in polynomial space, and $\pfp$ captures $\pspace$, then $\Phi$ can be expressed in $\pfp$. We define
$\alpha := \sa{X}(\Phi(X)\mult\varphi(X))$, where $\varphi(X) = \pa{Y}(\varphi_{<}(Y,X)\mapsto 2)$. Note that for each $B\subseteq A^{\ell}$ such that $B$ is the $m$-th element in the given order, $\sem{\varphi(B)}(\A) = 2^m$. Therefore, $\alpha\in\qso(\pfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$. \qed

\subsection{Proof of Theorem \ref{eqso-sigma-zero-in-fp}}

First, for a given signature $\R$ and logic fragment $\LL$ we define a {\em valid form in $\LL$} (or just {\em valid form}). A formula $\alpha$ is in valid form in $\LL$ if (1) it is equal to $\varphi$, $1$, $\sa{\bar{x}}\varphi$, $\sa{\bar{x}}1$, $\sa{\bar{X}}\varphi$, $\sa{\bar{X}}1$, $\sa{\bar{X}}\sa{\bar{x}}\varphi$ or $\sa{\bar{X}}\sa{\bar{x}}1$ for some $\LL$ formula $\varphi$ and some tuple of first-order (second-order) variables $\bar{x}$ ($\bar{X}$), or (2) if it is equal to $(\beta\add\gamma)$ where $\beta$ and $\gamma$ are formulas in valid form. Given a $\eqso(\loge{0})$ formula $\alpha$ over a signature $\R$ we will recursively construct a formula $\tau(\alpha)$ in valid form such that $\alpha(\enc(\A)) = \tau(\alpha)(\enc(\A))$ for each $\A\in\ostr[\R]$.

First, each natural constant $s$ in $\alpha$ is replaced by $(1\add 1\add\cdots\add 1)$ ($s$ times). If $\alpha = 1$ or $\alpha = \varphi \in \loge{0}$, then $\tau(\alpha) = \alpha$. We assume that every sub-formula of $\alpha$ is in valid form. If $\alpha = (\beta \add \gamma)$ then $\tau(\alpha) = \alpha$. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$ this case separates in more cases: If $\alpha = \sa{x}\beta$ and $\beta$ is in one of the valid forms in (1), then $\tau(\alpha) = \alpha$. If $\alpha = \sa{X}\beta$ and $\beta$ is in the valid form $\sa{\bar{x}}\gamma$ then $\tau(\alpha) = \sa{\bar{x}}\sa{X}\gamma$; if $\beta$ it is in one of the remaining valid forms in (1), then $\tau(\alpha) = \alpha$. If $\alpha = \sa{x}\beta$ ($\alpha = \sa{X}\beta$) and $\beta = (\gamma_1\add\gamma_2)$, then $\tau(\alpha) = \tau(\sa{x}\gamma_1) \add \tau(\sa{x}\gamma_2)$ ($\tau(\alpha) = \tau(\sa{X}\gamma_1) \add \tau(\sa{X}\gamma_2)$). This covers all possible cases for $\alpha$. Furthermore, suppose that $\LL$ is closed under conjunction and suppose $\alpha = (\beta\mult\gamma)$. If $\beta = (\delta_1\add\delta_2)$ (wlog), then $\tau(\alpha) = \tau(\delta_1\mult\gamma)\add\tau(\delta_2\mult\gamma)$. If $\beta = \sa{\bar{x}}\delta$, then $\tau(\alpha) = \tau(\sa{\bar{x}}\tau(\delta\mult\gamma))$. If $\beta = 1$, then $\tau(\alpha) = \gamma$. If $\beta = \varphi$ and $\gamma = \psi$, then $\tau(\alpha) = \varphi\wedge\psi$. Let $\alpha$ be a $\eqso(\loge{0})$ formula over a signature $\R$. Consider the following procedure that receives $\enc(\A)$ as input, where $\A\in\ostr[\R]$ with domain $A$. Let $\tau(\alpha) = \beta_1\add\cdots\add\beta_m$. For each $\beta_i$, if $\beta_i = 1$ or $\beta_i = \varphi$, then $\sem{\beta_i}(\A)$ can be computed in polynomial time. If $\beta_i = \sa{\bar{x}}1$ then $\sem{\beta_i}(\A) = \vert A \vert^{\length(\bar{x})}$, if $\beta_i = \sa{\bar{X}}1$ then $\sem{\beta_i}(\A) = 2^{\Sigma_{X\in\bar{X}}\vert A \vert^{\arity(X)}}$, if $\beta_i = \sa{\bar{X}}\sa{\bar{x}}1$ then $\sem{\beta_i}(\A) = \vert A \vert^{\length(\bar{x})}2^{\Sigma_{X\in\bar{X}}\vert A \vert^{\arity(X)}}$, all of which can be computed in polynomial time. Otherwise, $\beta_i$ is $\sa{\bar{x}}\varphi$ or $\sa{\bar{X}}\sa{\bar{x}}\varphi$, and then $\sem{\beta_i}(\A) = f_{\varphi}(\enc(\A))$, and $f_{\varphi}\in \#\loge{0}$ which implies that $f_{\varphi}(\enc(\A))$ can be computed in polynomial time \cite{SalujaST95}.

\subsection{Proof of Theorem \ref{eqso-sigma-one-in-eqso-pi-one}}

Let $\alpha$ be a formula in $\eqso(\loge{1})$ over a signature $\R$. Let $\tau(\alpha) = \beta_1\add\beta_2\add\cdots\add\beta_m$ defined as in the proof of Theorem \ref{eqso-sigma-zero-in-fp}. This is, each $\beta_i$ is of the form $1$, $\varphi$, $\sa{\bar{x}}1$, $\sa{\bar{x}}\varphi$, $\sa{\bar{X}}1$, $\sa{\bar{X}}\varphi$, $\sa{\bar{X}}\sa{\bar{x}}1$ or $\sa{\bar{X}}\sa{\bar{x}}\varphi$, where $\varphi$ is a formula in $\loge{1}$ and $\bar{x}$ ($\bar{X}$) is a tuple of first-order (second-order) variables. For each $\beta_i$ we will construct a $\gamma_i$ such that $\gamma_i \in \eqso(\logu{1})$ and $\sem{\beta_i}(\A) = \sem{\gamma_i}(\A)$ for each $\A\in\ostr[\R]$. Since the forms $1$, $\sa{\bar{x}}1$, $\sa{\bar{X}}1$ and $\sa{\bar{X}}\sa{\bar{x}}1$ are already valid in $\eqso(\logu{1})$, we suppose $\beta_i$ is in one of the remaining forms. Without loss of generality, let $\beta_i = \sa{\bar{X}}\sa{\bar{x}}\exists\bar{y}\,\varphi(\bar{x},\bar{y},\bar{X})$ where $\varphi(\bar{x},\bar{y},\bar{X})$ is a quantifier-free formula. Similarly as in the proof by Saluja et al. \cite{SalujaST95}, we define $\gamma_i = \sa{\bar{x}}\sa{\bar{y}}\sa{\bar{X}}\varphi(\bar{x},\bar{y},\bar{X}) \wedge \forall\bar{z}(\varphi(\bar{x},\bar{z},\bar{X})\to(\bar{z} < \bar{y} \vee \bar{z} = \bar{y}))$ that fulfils the desired condition. The theorem holds true by using $\beta = \gamma_1\add\cdots\add\gamma_m$ which is in $\eqso(\logu{1})$

%\subsection{Proof of Theorem \ref{one-sigma-zero}}
%
%Towards a contradiction, suppose that there exists a constant function $f(\A) = s > 0$ which is expressible in $\E{0}$. This is, there is a formula $\psi(x,X)$ such that $f_{\psi(x,X)}(\A) = \vert \{ \langle e,P\rangle \mid \A\models\psi(e,P) \} \vert = s$ for each $\A\in\ostr$. It can be seen that if $X$ contains more than zero variables, then the function $f_{\psi(x,X)}$ grows exponentially. We have now that $\psi(x)$ contains only first order variables. Consider some $\A \in \ostr$ of size $n$ and let $a_1,\ldots,a_s\in \A$ be the tuples that satisfy $\A \models \psi(a_i)$. Now consider a structure $\A'$ that is obtained by duplicating $\A$. This is, $\A'$ has size $2n$, its first half is isomorphic to its second half, and also to $\A$. We have that there is at least $2s$ tuples $a_i,b_i\in \A'$, $s$ on each half, such that $\A'\models\psi(a_i)$ and $\A'\models\psi(b_i)$, which leads to a contradiction.

%\textbf{Proof of Theorem \ref{mult-sigma-zero}}
%
%Let $\alpha \in \sqso(\loge{0})$ with binary product. This is, $\alpha$ is defined by the grammar $\alpha := \sa{x}\varphi \ \mid \ \sa{X}\varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \ (\alpha\mult\alpha)$, where $\varphi \in \loge{0}$. For convenience, we add the expressions $\sa{x}(\varphi\mult\varphi)$ and $\sa{X}(\varphi\mult\varphi)$ to the grammar. Let $k$ be the number of algebraic operators in $\alpha$. We will show by induction on $k$ that $\alpha$ is expressible without multiplication. If $k = 1$, then the possible cases are $\alpha = \sa{x}\varphi$, where $\varphi\in\loge{0}$, which is already product-free. If $k = 2$, then the only that is not two nested summations is $\alpha = \sa{x}(\varphi\mult\psi)$, which can be expressed as $\alpha = \sa{x}(\varphi\wedge\psi)$. Suppose that the hypothesis holds for $k = 1$ through $n-1$. Let $\alpha$ have $n$ algebraic operators. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$ where $\beta$ is an algebraic expression, then $\beta$ has $n-1$ algebraic operators and by hypothesis it can be expressed as product-free. If $\alpha = (\beta \mult \gamma)$ there are two cases. If at least one between $\beta$ and $\gamma$ has a product, then by hypothesis it can be expressed as product-free. Moreover, the new expression for $\alpha$ has strictly less than $n$ algebraic operators so by hypothesis it can be expressed as product-free. If there are no products in $\beta$ or $\gamma$, then let $\beta = \sa{x}\delta(x)$. We have that $\alpha$ can be expressed as $\sa{y}(\delta(y)\mult\gamma)$, where $y$ is a first-order variable that does not appear in $\beta$ or $\gamma$. Since $\delta(y)\mult\gamma$ has $n-1$ algebraic operators, it can be expressed as product-free, and then $\alpha$ can be expressed as product-free.
%
%\textbf{Proof of Theorem \ref{fo-prod-sigma-zero}}
%
%The resulting grammar is $\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \pa{x}\alpha$. Consider a more restricted grammar that replaces $\pa{x}\alpha$ by $\pa{x}\sa{y}\varphi$. Let $\alpha = \pa{x}\sa{y}(x = y)$. This formula produces a constant function equal to 1, which as shown in Theorem \ref{one-sigma-zero} is not expressible in $\E{0}$. 
%
%\textbf{Proof of Theorem \ref{fp1}}
%
%Let $f \in \XE{0}$, and let $\varphi(\X,\x)$ be and {\sc FO}-extended quantifier-free $\L$-formula such that:
%\begin{eqnarray*}
%	f(\A) &=& \vert \{\langle \P,\e  \rangle \mid \A \models \varphi(\P,\e) \} \vert
%\end{eqnarray*}
%for each $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \str$, where $\e \in A^m$ and $\P = (P_1,\ldots,P_q)$ is a tuple of predicates. We will now show that computing $f(\A)$ can be done in polynomial time.
%
%For each {\sc FO}-formula $\beta(\x)$ in $\varphi(\X,\x)$, let $R_{\beta}$ be a predicate of arity $m$. Let $\A^\prime = \langle A, \S^{\A}, R_{\beta}^{\A^\prime}, \leq^{\A} \rangle \in \str$, where $R_{\beta}^{\A^\prime} = \{\d \mid \A \models \beta(\d)\}$. Note that each $\beta(\x)$ is fixed in $\varphi(\X,\x)$, and for each $\d \in A^m$, checking whether $\A \models \beta(\d)$ can be done in polynomial time. Therefore, generating $R_{\beta}^{\A^\prime}$ can also be done in polynomial time.
%
%Let $\psi(\X,\x)$ be obtained by replacing each {\sc FO}-formula $\beta(\x)$ in $\varphi(\X,\x)$ by $R_{\beta}(\x)$. Also, let $g = f_{\psi(\X,\x)}$. Note that for each tuple of predicates $\P$ and each $\e \in A^m$, $\A \models \varphi(\P,\e)$ if and only if $\A^\prime \models \psi(\P,\e)$, and so, $g(\A^\prime) = f(\A)$. But, $\psi(\X,\x)$ is a quantifier-free $\L$-formula, and therefore, $g \in \#\Sigma_0$. Since it is shown in \cite{DBLP:journals/jcss/SalujaST95} that $\#\Sigma_0 \subseteq$ {\sc FP}, we conclude that $f(\A)$ can be evaluated in polynomial time.

\textbf{Proof of Theorem \ref{decisionptime}}

Let $f$ be a function in $\E{1}$. Then there is an extended quantifier-free $\L-$formula $\varphi(\x,\X,\z)$ such that
\begin{eqnarray*}
	f(\A) &=& \mid \{ \langle\P,\e\rangle \mid \A \models \exists \x \, \varphi(\x,\P,\e) \} \mid,
\end{eqnarray*}
where $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \str$, $A = \{a_1,\ldots,a_{\vert A \vert}\}$, $\z$ is an $m$-tuple of variables and $\x$ is a $k$-tuple of variables. Let $\y = (\x,\z)$ and let $\psi(\X,\y) = \varphi(\x,\X,\z)$. Moreover, let $g = f_{\psi(\X,\y)}$.
\begin{claim} \label{mono}
	For each $\A \in \str$, $f(\A) > 0$ iff $g(\A) > 0$.
\end{claim}
\begin{proof}
	($\Rightarrow$) Suppose $f(\A) > 0$. Let $\P$ and $\e$ be such that $\A \models \exists \x \, \varphi(\x,\P,\e)$. It follows that there is at least one $\d \in A^k$ such that $\A \models \varphi(\d,\P,\e) = \psi(\P,(\d,\e))$. Therefore, $g(\A) > 0$.

	($\Leftarrow$) Suppose $g(\A) > 0$. Let $\Q$ and $\c = (\c_1,\c_2)$, where $\c_1$ and $\c_2$ have $k$ and $m$ elements respectively, be such that $\A \models \psi(\Q,\c) = \varphi(\c_1,\Q,\c_2)$. Then we have that $\A \models \exists \x \, \varphi(\x,\Q,\c_2)$, from which we conclude that $f(\A) > 0$.
\end{proof}
Note that $\psi(\X,\y)$ is an {\sc FO}-extended $\L$-formula, so $g \in\XE{0}$. It follows from \ref{mono} that for each $\A \in \str$, it holds that $\A \in L_f$ if and only if $g(\A) > 0$. But by Theorem \ref{fp1}, $g(\A)$ can be evaluated in polynomial time. From this we conclude that the decision version of $f$ is in $\ptime$.

\textbf{Proof of Theorem \ref{sub-pnp}}

Suppose that $\E{1}$ is closed under substraction, that is, for each pair of functions $f,g\in \E{1}$, there exists $h\in\E{1}$ such that $(f\dotminus g)(\A) = h(\A)$ for each $\A\in\str$.

>>>>

@@

If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$. We also denote $\length(\bar{x}) = m$ ($\length(\bar{X}) = m$).
